How functional traits influence plant growth and shade tolerance across the life-cycle

### 2018-04-05 ms/ms.tex
Enable line numbers, in grey
Revise paper after reviewers comments

\documentclass[9pt,twocolumn,twoside]{pnas-new}

@@

\usepackage{graphicx}
\graphicspath{{../output/}}

\newcommand{\smurl}[1]{\url{#1}}
\newcommand{\ud}{\ensuremath{\rm{d}}}
\newcommand{\tabitem}{~~\llap{\textbullet}~~}
\newcommand{\email}[1]{\href{mailto:#1}{\texttt{#1}}}
\newcommand{\plant}{\texttt{plant}}
\newcommand{\wplcp}{\textsc{wplcp}}
\newcommand{\sma}{\textsc{sma}}

@@

\title{Trajectories: how functional traits influence plant growth and shade tolerance across the life-cycle}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Daniel S. Falster}
\author[b]{Remko A. Duursma}
\author[a,c]{Richard G. FitzJohn}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Evolution \& Ecology Research Centre, School of Biological, Earth and
Environmental Sciences, University of New South Wales, Sydney, NSW 2052, Australia}
\affil[b]{Hawkesbury Institute for the Environment, Western Sydney University, Locked Bag 1797, Penrith NSW 2751, Australia}
\affil[c]{Imperial College, London, United Kingdom}

% Please give the surname of the lead author for the running footer
\leadauthor{Falster}

% Please add here a significance statement to explain the relevance of your work
% 120 words
% Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. The Significance Statement appears in the paper itself and is required for all research papers. Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. The Significance Statement appears in the paper itself.
\significancestatement{
Plant species differ in many functional traits -- features of specific tissues and allocation of energy among them. While traits have been used in many correlative approaches to describe communities and demography, it has remained unclear how and why traits should influence whole-plant growth. Here we present a new theoretical framework for understanding the effect of traits on plant growth and shade-tolerance. This framework captures diverse patterns of growth in relation to size, and explains why the effect of traits on growth changes through ontogeny. By disentangling the effects of plant size, light environment and traits on growth rates, this study provides a theoretical foundation for understanding growth across diverse tree species around the world.
}

@@

% 250 words
\begin{abstract}
Plant species differ in many functional traits that drive differences in rates of photosynthesis, biomass allocation, and tissue turnover. Yet, it remains unclear how -- and even if -- such traits influence whole-plant growth, with the simple linear relationships predicted by existing theory often lacking empirical support. Here we present a new theoretical framework for understanding the effect of diverse functional traits on plant growth and shade-tolerance by extending a widely-used model, linking growth rate in seedlings with a single leaf trait, to explicitly include influences of size, light environment, and five other prominent traits: seed mass, height at maturation, leaf mass per unit leaf area, leaf nitrogen per unit leaf area, and wood density. Based on biomass production and allocation, this framework explains why the influence of traits on growth rate and shade tolerance often varies with plant size and why the impact of size on growth varies among traits. Specifically, we demonstrate why for height growth the influence of: i) leaf mass per unit leaf area is strong in small plants but weakens with size, ii) leaf nitrogen per unit leaf area does not change with size, iii) wood density is present across sizes, iv) height at maturation strengthens with size, and v) seed mass decreases with size. Moreover, we show how traits moderate plant responses to light environment and also determine shade tolerance, supporting diverse empirical results.
\end{abstract}

@@

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalized the article contents.
\verticaladjustment{-2pt}

@@

\dropcap{F}unctional traits capture core differences in the strategies plants use to generate and invest resources \citep{Westoby-2002, Wright-2004, Chave-2009}. Although most woody plants have the same basic physiological function and key resource requirements (carbon, nutrients and water), species differ considerably in the rates at which resources are acquired, invested into different tissues, and lost via turnover. During the last two decades, trade-offs related to some prominent traits have been quantified, with values for traits such as leaf mass per unit leaf area, wood density, and seed size now available for up to 10\% of the world's 250000 plant species \citep{Cornwell-2014}. As data has accumulated, researchers are increasingly looking to traits to predict patterns in plant growth, demography, life history and performance \citep{Poorter-2008, Wright-2010, VanKleunen-2010, Adler-2014}. In this article we outline the mechanisms by which the growth of individual plants is influenced by various functional traits, as well as plant size and the light environment.

While the influence of traits on elements of plant physiological function has been increasingly quantified and understood, attempts at using traits to predict demographic rates have met with mixed success \citep{Poorter-2006, Poorter-2008,Wright-2010,Herault-2011,Paine-2015}. In seedlings, the trait leaf mass per unit leaf area ($\phi$) -- the central element of the leaf economics spectrum \citep{Wright-2004} -- was found to be tightly correlated with relative growth rate in plant mass \citep{Lambers-1992, Cornelissen-1996, Wright-2000}, as predicted from a simple mathematical model of growth rate (see below). $\phi$ and leaf lifespan, itself closely correlated with $\phi$, were also linked to height growth rate for small seedlings and saplings \citep{Reich-1992, Poorter-2006}. These early successes prompted researchers to search for similar relationships in large plants. However, the results showed that in saplings and trees, $\phi$ was not correlated with growth rate \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015}. Meanwhile, other traits such as wood density showed strong relationships to growth in large plants \citep{Wright-2010,Herault-2011}, but less so in small plants \citep{Castro-1998}. Clearly, traits do not correlate simply and consistently to growth rate.

Recently, it has become clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations provides the most compelling evidence \citep{Gibert-2016}. That study showed that the strength of the correlation between some traits (including $\phi$ and maximum height) and growth rate was modified by size, while for other traits (including wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Here, we extend previous work \citep{Gibert-2016} to provide a mechanistic explanation of these empirical patterns via a model of plant growth and allocation.

\begin{table}[!ht]
\caption{Empirical phenomena explained in this paper.}
\begin{tabular}{p{8cm}}
\toprule
\textbf{Change in growth rate with increasing size (Fig. \ref{fig:conceptual})}\\
 $\,$ Net biomass production: Hump-shaped \citep{Givnish-1988, Koch-2004} \\
 $\,$ Plant mass: Increasing \citep{Sillett-2010, Stephenson-2014} \\
 $\,$ Height: Hump-shaped \citep{Ryan-2006, Sillett-2010, King-2011} \\
 $\,$ Stem-diameter: Hump-shaped \citep{Canham-2004, Canham-2006, Herault-2011} \\
 $\,$ Relative growth rate (all variables): decreasing \citep{Rees-2010, Iida-2014}\\
\textbf{Effect of traits on growth rate (Fig. \ref{fig:growth_light_height})}\\
 $\,$ $\omega$: low values $\rightarrow$ smaller seedlings, lower absolute \& higher relative growth \citep{Gibert-2016} \\
 $\,$ $H_{\rm m}$: low values $\rightarrow$ slower growth at at larger sizes \citep{Gibert-2016}\\
 $\,$  $\nu$: low values $\uparrow$ growth irrespective of size, but only in high light \citep{Gibert-2016}\\
 $\,$ $\phi$: low values $\uparrow$ growth when small, not at mid-large sizes \citep{Gibert-2016}\\
 $\,$ $\rho$: low values $\uparrow$ growth, except at largest sizes \citep{Gibert-2016}\\
\textbf{Responsiveness of growth rate to changes in light, $E$ (Fig. \ref{fig:growth_light_height})}\\
 $\,$ $\nu$: high values respond more \\
 $\,$ $\phi$: low values respond more \\
 $\,$ $\rho$: low values respond more \citep{Ruger-2012}\\
\textbf{Shade tolerance, \textsc{wplcp} (Fig. \ref{fig:wplcp})}\\
 $\,$ Decreasing with size \citep{Givnish-1988, Kneeshaw-2006, Lusk-2008}\\
 $\,$ $\nu$: low values $\uparrow$ shade tolerant \citep{Messier-1999, Craine-2005,Baltzer-2007}\\
 $\,$ $\phi$: low values $\downarrow$ shade tolerant \citep{Messier-1999, Poorter-2006, Baltzer-2007, Lusk-2008}$^3$, higher \textsc{lai} \citep{Reich-1992, Gower-1993, Niinemets-2010} \\
 $\,$ $\rho$: low values less shade tolerant \citep{Osunkoya-1996}\\
  \bottomrule
\end{tabular}
\addtabletext{Traits are seed mass ($\omega$), height at maturation ($H_{{\rm mat}}$), leaf nitrogen content per unit leaf area ($\nu$), leaf mass per unit leaf area ($\phi$), and wood density ($\rho$); for details see Tables \ref{tab:definitions}-\ref{tab:traits}. \textsc{lai} = leaf area index. Footnotes: $^1$ Similar responses are predicted for maximum photosynthetic rate per leaf area \& dark respiration rate per leaf area; here both directly related to $\nu$. $^2$ Similar responses are predicted for leaf lifespan; here directly related to $\phi$. $^3$ \citep{Baltzer-2007} reports a strong relationship between \textsc{wplcp} and leaf respiration rate per unit leaf area or mass.

@@

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{main/figure_conceptual.pdf}
\caption{Conceptual framework linking growth rate to plant size and traits.
\textbf{a)} Shows how the distribution of mass in a typical plant varies with size.
\textbf{b)} Equations describing the rates of biomass production and growth in various dimensions of the plant. In the first line the symbol $\Sigma$ means ``sum'' across tissues, where $i = {\rm l,b,s,r}$. Grey numbers indicate equation numbers referred to in the main text. The insets show how the different metrics change intrinsically with plant height, when applying the ``functional-balance'' model in Table \ref{tab:allometry}. Colours highlight where the same metric appears repeatedly in different equations. For a list of variable names see Table \ref{tab:definitions}.}
\label{fig:conceptual}
\end{figure}

Interpreting diverse empirical results seeking to link traits to growth rate is challenging because, until recently \citep{Gibert-2016}, we lacked any clear expectations on why the effect of traits may be moderated by size. A widely-used equation for seedlings suggests that, all else being equal, a seedling's relative growth rate in mass is linearly and negatively related to $\phi$ \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. An extension of this equation suggests a similar relationship should hold at larger plant sizes \citep{Enquist-2007}. But as noted above, the prediction for large plants is not supported by empirical results. Meanwhile, theoretical predictions on how other traits should influence growth are largely absent. Without further guidance, many researchers have expected traits to map directy to growth rates, i.e. there is a ``fast'' and ``slow'' growth traits \citep[e.g.][]{Grime-1977, Poorter-2008, Chave-2009, Paine-2015}.

One reason theoretical predictions have been lacking or not been supported is that in existing theory the effects of traits is realised mainly via influences on primary productivity (photosynthesis - respiration) \citep{Wright-2000, Enquist-2007}. By contrast, the physiology of traits such as $\phi$ and wood density suggests they influence allocation of biomass among different tissues rather than mass production \citep{Falster-2011, Duursma-2016, Gibert-2016}. A second concern for theory focussed on mass production is that measuring mass production is really only practical for small plants. On larger plants, growth is measured mainly as increment in either stem diameter or height \citep{Purves-2008, Anderson-2015, Kunstler-2016}. As we outline below, height and diameter growth depend not only on mass production but also allocation of biomass and tissue construction costs, meaning theory considering only primary production of limited utility.

\begin{figure}[!ht]
\centering
\includegraphics[width=.9\linewidth]{main/lcp_schematic.pdf}
\caption{Conceptual framework linking shade-tolerance to plant size and traits, adapted from \citep{Givnish-1988}. Shade tolerance is the level of canopy openness $E$ where photosynthetic income (dashed line) intersects with the sum of respiration and turnover costs over tissues (solid black lines). The black circles indicate the point of intersection for plants with different heights. All income and cost is expressed in units of dry mass produced  per unit leaf area per year. Note the costs of sapwood and bark increases with height.
Traits can impact on shade tolerance by, in the case of leaf nitrogen content, shifting the income line up or down, or in the case of leaf mass per unit leaf area or wood density, causing the cost components to increase or decrease.
\label{fig:wplcp_idea}}
\end{figure}

\begin{table}[!hb]
 \caption{Variable definitions.}
\centering
\begin{tabular}{p{0.5cm}p{1.75cm}p{5cm}}
\toprule
Symbol & Unit & Description \\
\midrule
\multicolumn{3}{l}{\textbf{Traits (constant through ontogeny)}}\\
$\omega$ & kg & Seed mass  \\
$H_{{\rm mat}}$ & m & Height at maturation\\
$\nu$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Leaf nitrogen per unit leaf area  \\
$\phi$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Leaf mass per area \\
$\rho$ & $\mathrm{kg}\,\mathrm{m}^{-3}$ & Wood density \\
\multicolumn{3}{l}{\textbf{State variables (may vary through ontogeny)}} \\
$H$ & m & Height of a plant\\
$B$ & kg & Biomass originating from parent plant\\
$D$ & m & Stem diameter\\
$M_i$ & kg & Mass of tissue type $i$ retained on plant \\
$A_i$ & m$^2$ & Surface or cross-section are of tissue type $i$\\
$p,\bar{p}$ & mol yr$^{-1}$ m$^{-2}$ & Photosynthetic rate per unit area \\
$r_i$ & mol yr$^{-1}$ kg$^{-1}$  & Respiration rate per unit mass of tissue type $i$ \\
$k_i$ & yr$^{-1}$ & Turnover rate for tissue type $i$ \\
\multicolumn{3}{l}{\textbf{Environmental variables (fixed)}} \\
$E$ & & Canopy openness\\
\multicolumn{3}{l}{\textbf{Other parameters (constant throughout)}} \\
$\alpha_{\rm y}$ &  & Yield, fraction of carbon fixed into biomass\\
$\alpha_{\rm bio}$  & $\mathrm{kg}\,\mathrm{mol}^{-1}$ & Biomass per mol carbon \\
$\eta_c$ & & Crown-shape parameter\\
$\theta$ &  & Sapwood area per unit leaf area\\
$\alpha_{\rm l1}$ & m & Height of plant with leaf area of 1m$^2$ \\
$\alpha_{\rm r1}$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Root mass per unit leaf area \\
$\alpha_{\rm b1}$ &  & Ratio of bark area to sapwood area\\
\bottomrule
\end{tabular}
\addtabletext{For $M_i$, $r_i$, $k_i$ and \(A_i\), subscripts refer to: \({\rm l}\) = leaves, \({\rm b}\) = bark, \({\rm s}\) = sapwood,  \({\rm h}\) = heartwood, \({\rm r}\) = roots, \({\rm st}\) = total stem, \({\rm a}\) =  alive tissue.

@@

\matmethods{
\vspace{-1em}
\subsection{Trait-based trade-offs}
The trade-offs described in Table \ref{tab:traits} were embedded within the growth model as follows.

We assume a direct trade-off between a plant's fecundity and $\omega$, with larger seeds resulting in larger seedlings.

>>>>

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}

@@

\input{ms-common-defs}

@@

\title{How functional traits influence plant growth and shade tolerance across the life-cycle}

% Please add here a significance statement to explain the relevance of your work
% 120 words
% Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. The Significance Statement appears in the paper itself and is required for all research papers. Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. The Significance Statement appears in the paper itself.
\significancestatement{
Plant species differ in many functional traits -- features of specific tissues and allocation of energy among them. While traits have been used in many correlative approaches to describe communities and demography, it has remained unclear how and why traits should influence whole-plant growth. Here we present a new theoretical framework for understanding the effect of traits on plant growth and shade-tolerance. This framework captures diverse patterns of growth in relation to size, and explains why the effect of traits on growth changes through ontogeny. By disentangling the effects of plant size, light environment, and traits on growth rates, this study provides a theoretical foundation for understanding growth across diverse tree species around the world.
}

@@

% 250 words
\begin{abstract}
Plant species differ in many functional traits that drive differences in rates of photosynthesis, biomass allocation, and tissue turnover. Yet, it remains unclear how -- and even if -- such traits influence whole-plant growth, with the simple linear relationships predicted by existing theory often lacking empirical support. Here we present a new theoretical framework for understanding the effect of diverse functional traits on plant growth and shade-tolerance by extending a widely-used model, linking growth rate in seedlings with a single leaf trait, to explicitly include influences of size, light environment, and five other prominent traits: seed mass, height at maturation, leaf mass per unit leaf area, leaf nitrogen per unit leaf area, and wood density. Based on biomass growth and allocation, this framework explains why the influence of traits on growth rate and shade tolerance often varies with plant size and why the impact of size on growth varies among traits. Specifically, we demonstrate why for height growth the influence of: i) leaf mass per unit leaf area is strong in small plants but weakens with size, ii) leaf nitrogen per unit leaf area does not change with size, iii) wood density is present across sizes, iv) height at maturation strengthens with size, and v) seed mass decreases with size. Moreover, we show how traits moderate plant responses to light environment and also determine shade tolerance, supporting diverse empirical results.
\end{abstract}

@@

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\begin{document}
% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalized the article contents.
\verticaladjustment{-2pt}

@@

\dropcap{F}unctional traits capture core differences in the strategies plants use to acquire and invest resources \citep{Westoby-2002, Wright-2004, Chave-2009}. Although most woody plants have the same basic physiological function and key resource requirements (carbon, nutrients, and water), species differ considerably in the rates at which resources are acquired, invested into different tissues, and lost via turnover. During the last two decades, trade-offs related to some prominent traits have been quantified, with values for some leaf, wood, or seed traits now available for up to 10\% of the world's 250000 plant species \citep{Cornwell-2014}. As data has accumulated, researchers are increasingly looking to traits to predict patterns in plant growth, demography, life history and performance \citep{Poorter-2008, Wright-2010, VanKleunen-2010, Adler-2014}. In this article we outline the mechanisms by which the growth of individual plants is influenced by various functional traits, as well as plant size and the light environment.

While the influence of traits on elements of plant physiological function has been increasingly quantified and understood, attempts at using traits to predict demographic rates have met with mixed success \citep{Poorter-2006, Poorter-2008,Wright-2010,Herault-2011,Paine-2015}. In seedlings, the trait leaf mass per unit leaf area ({\lma}) -- the central element of the leaf economics spectrum \citep{Wright-2004} -- was found to be tightly correlated with relative growth rate in plant mass \citep{Lambers-1992, Cornelissen-1996, Wright-2000}, as predicted from a simple mathematical model of growth rate (see below). {\lma} and a close correlate leaf lifespan, were also linked to height growth rate for small seedlings and saplings \citep{Reich-1992, Poorter-2006}. These early successes prompted researchers to search for similar relationships in large plants. However, in saplings and trees showed, {\lma} was not correlated with growth rate \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015}. Meanwhile, other traits such as wood density ({\wood}) showed strong relationships to growth in large plants \citep{Wright-2010,Herault-2011}, but less so in small plants \citep{Castro-1998}. Clearly, traits do not correlate simply and consistently to growth rate.

Recently, it has become clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations provides the most compelling evidence \citep{Gibert-2016}. That study showed that the strength of the correlation between some traits (including {\lma} and maximum height) and growth rate was modified by size, while for other traits (including {\wood} and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Here, we extend previous work \citep{Gibert-2016} to provide a mechanistic explanation of these empirical patterns via a model of plant growth and allocation.

\begin{table}[!ht]
\caption{Empirical phenomena explained in this paper.}
\begin{tabular}{p{8cm}}
\toprule
\textbf{Change in growth rate with increasing size (Fig. \ref{fig:conceptual})}\\
 $\,$ Biomass growth: Hump-shaped \citep{Givnish-1988, Koch-2004} \\
 $\,$ Plant mass: Increasing \citep{Sillett-2010, Stephenson-2014} \\
 $\,$ Height: Hump-shaped \citep{Ryan-2006, Sillett-2010, King-2011} \\
 $\,$ Stem-diameter: Hump-shaped \citep{Canham-2004, Canham-2006, Herault-2011} \\
 $\,$ Relative growth rate (all variables): decreasing \citep{Rees-2010, Iida-2014}\\
\textbf{Effect of traits on growth rate (Fig. \ref{fig:growth_light_height})}\\
 $\,$ {\seed}: $\downarrow$ values $\downarrow$ seedling size, \& thus $\downarrow$ absolute \& $\uparrow$ relative growth rate \citep{Gibert-2016} \\
 $\,$ $H_{\rm mat}$: $\downarrow$ values $\downarrow$ growth rate at at larger sizes \citep{Gibert-2016}\\
 $\,$  {\nitrogen}: $\downarrow$ values $\uparrow$ growth rate irrespective of size, but only in high light \citep{Gibert-2016}\\
 $\,$ {\lma}: $\downarrow$ values $\uparrow$ growth rate when small, not at mid-large sizes \citep{Gibert-2016}\\
 $\,$ {\wood}: $\downarrow$ values $\uparrow$ growth rate, except at largest sizes \citep{Gibert-2016}\\
\textbf{Responsiveness of growth rate to changes in light, $E$ (Fig. \ref{fig:growth_light_height})}\\
 $\,$ {\nitrogen}: $\uparrow$ values respond more \\
 $\,$ {\lma}: $\downarrow$ values respond more \\
 $\,$ {\wood}: $\downarrow$ values respond more \citep{Ruger-2012}\\
\textbf{Shade tolerance, \textsc{wplcp} (Fig. \ref{fig:wplcp})}\\
 $\,$ Size: decreasing \citep{Givnish-1988, Kneeshaw-2006, Lusk-2008}\\
 $\,$ {\nitrogen}: $\downarrow$ values $\uparrow$ shade tolerance \citep{Messier-1999, Craine-2005,Baltzer-2007}\\
 $\,$ {\lma}: $\downarrow$ values $\downarrow$ shade tolerance \citep{Messier-1999, Poorter-2006, Baltzer-2007, Lusk-2008}$^3$ \& $\uparrow$ \textsc{lai} \citep{Reich-1992, Gower-1993, Niinemets-2010} \\
 $\,$ {\wood}: $\downarrow$ values $\downarrow$ shade tolerance \citep{Osunkoya-1996}\\
  \bottomrule
\end{tabular}
\addtabletext{{\seed} = seed mass, {\hmat}= height at maturation, {\nitrogen}= leaf nitrogen content per unit leaf area, {\lma}=leaf mass per unit leaf area, {\wood}=wood density, \textsc{lai} = leaf area index. Footnotes: $^1$ Similar responses are predicted for maximum photosynthetic rate per leaf area \& dark respiration rate per leaf area; here both related to {\nitrogen}. $^2$ Similar responses are predicted for leaf lifespan; here directly related to {\lma}. $^3$ \citep{Baltzer-2007} finds a relationship between \textsc{wplcp} \& leaf respiration rate expressed per unit leaf area or per unit leaf mass.

@@

\begin{table}[!hb]
 \caption{Key trade-offs (benefit and cost) for the five traits considered, as encoded into the {\plant} model.
 }
\centering
  \begin{tabular}{p{0.5cm}p{3.6cm}p{2.5cm}p{0.4cm}}
  \toprule
  Trait &  Benefit & Cost & Ref. \\
  \midrule
  {\seed} & $\uparrow$ size of seedlings & $\downarrow$ seed production rate & \citep{Moles-2006}\\
  {\hmat}  & $\uparrow$ growth rate & $\downarrow$ seed production rate & \citep{Thomas-2011}\\
  {\nitrogen} & $\uparrow$ photosynthetic rate in high light & $\uparrow$ respiration rate & \citep{Wright-2004}\\
  {\lma} & $\downarrow$ leaf turnover rate & $\uparrow$ cost of building leaf & \citep{Wright-2004}\\
  {\wood} & $\downarrow$ sapwood turnover rate & $\uparrow$ cost of building stem & \citep{Chave-2009}\\
  \bottomrule
  \end{tabular}
\addtabletext{Traits: {\seed} = seed mass, {\hmat}= height at maturation, {\nitrogen}= leaf nitrogen content per unit leaf area, {\lma}=leaf mass per unit leaf area, {\wood}=wood density.

}
\label{tab:traits}
\end{table}

Interpreting diverse empirical results seeking to link traits to growth rate has been challenging because, until recently \citep{Gibert-2016}, we lacked any clear expectations on why the effect of traits may be moderated by size. A widely-used equation for seedlings suggests that, all else being equal, a seedling's relative growth rate in mass is linearly and negatively related to {\lma} \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. An extension of this equation suggests a similar relationship should hold at larger plant sizes \citep{Enquist-2007}. But as noted above, the prediction for large plants is not supported by empirical results. Meanwhile, theoretical predictions on how other traits should influence growth are largely absent. Without further guidance, many researchers expected traits to map directly onto growth rates \citep[e.g.][]{Grime-1977, Poorter-2008, Chave-2009, Paine-2015}.

One reason theoretical predictions have been lacking or not been supported is that in existing theory the effects of traits is realised mainly via influences on net primary productivity (photosynthesis - respiration) \citep{Wright-2000, Enquist-2007}. By contrast, the physiology of traits such as {\lma} and {\wood} suggests they influence allocation of biomass among different tissues, and their turnover rates, rather than net primary production \citep{Falster-2011, Duursma-2016, Gibert-2016}. A second concern for theory focussed on net primary production is that measuring production is really only practical for small plants. On larger plants, growth is measured mainly as increment in either stem diameter or height \citep{Purves-2008, Anderson-2015, Kunstler-2016}. As outlined below, growth in height and diameter also depend on the allocation of biomass and tissue construction costs.

Here we show how a new mechanistic growth model -- called {\plant} \citep{Falster-2016} -- can explain diverse empirical phenomena, including a size-dependent effect of traits on growth and an effect of traits on shade tolerance (Table \ref{tab:phenomena}), and thereby offer new insights into the way traits influence key elements of plant demography across the life-cycle. The {\plant} model builds on past approaches to modelling production and allocation of biomass \citep[e.g.][]{Givnish-1988, Yokozawa-1995,Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011, Gibert-2016}. Our primary focus in this article is to explain a pattern that has been gradually emerging -- that the effect of traits on plant growth is modified by plant size \citep{Ruger-2012, Iida-2014, Gibert-2016}. Based on the same decomposition of growth rates as is implemented in the {\plant} model and used below, \citep{Gibert-2016} argued conceptually why the effect of traits on growth should change with size. Here we extend the results of \citep{Gibert-2016} to provide a full, functioning model and use this to show, from the point of view of mass production and allocation, how and why the effect of some traits on growth rates changes with size. We consider five prominent traits representing specific physiological trade-offs: seed mass ({\seed}), height at maturation ({\hmat}), leaf nitrogen content per unit leaf area ({\nitrogen}),  {\lma}, and {\wood} (Table \ref{tab:traits}).

We also show how our approach can account for related phenomena, including changes in growth and shade tolerance with traits, individual size, and light environment (Table \ref{tab:phenomena}). Our view is that trait-based approaches -- which aim to explain differences among species -- should be integrated within a general model of plant growth, and thus should also be able to capture patterns of growth through ontogeny. In absolute terms, growth rates tend to show hump-shaped relationship with size, when expressed as either height \citep{Sillett-2010, King-2011}, diameter \citep{Canham-2004, Canham-2006, Herault-2011} or biomass growth \citep{Givnish-1988, Koch-2004}. In contrast, the growth rate of standing plant mass continues to increase with size \citep{Sillett-2010, Stephenson-2014}. Expressed as relative growth rates, all growth measures decrease sharply with size \citep{Rees-2010, Iida-2014}. Shade tolerance also varies among species, correlates with traits \citep{Messier-1999, Lusk-2008, Poorter-2006}, and tends to decrease with increasing size \citep{Givnish-1988, Kneeshaw-2006, Lusk-2008}. These diverse phenomena deserve an integrated explanation.

\begin{figure}[!ht]
\centering
\includegraphics[width=\linewidth]{main/figure_conceptual.pdf}
\caption{Conceptual framework linking growth rate to plant size and traits.
\textbf{a)} Shows how the distribution of mass in a typical plant varies with size.
\textbf{b)} Equations describing the rates of biomass growth, and growth in various dimensions of the plant. In the first line the symbol $\Sigma$ means ``sum'' across tissues, where $i = {\rm l,b,s,r}$. Grey numbers indicate equation numbers referred to in the main text. The insets show how the different metrics change intrinsically with plant height, when applying the ``functional-balance'' model in Table \ref{tab:allometry}. Colours highlight where the same metric appears repeatedly in different equations. For a list of variable names see Table \ref{tab:definitions}.}
\label{fig:conceptual}
\end{figure}

\section*{Modeling the effects of traits on demography}

The {\plant} model builds on the widespread approach used in many vegetation models of explicitly modelling the amounts of biomass in different tissues within a plant \citep[e.g.][]{Givnish-1988, Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011, DeKauwe-2014} (Fig. \ref{fig:conceptual}a). We consider the masses $M_i$, areas $A_i$, and diameters $D_i$ of tissues, where the subscripts indicates tissue type: l=leaf, b=bark and phloem, s=sapwood, h=heartwood, r=root, a=alive, t=total, and st=stem total (s+b+h). The total mass of living tissue is then $M_{\rm a}=M_{\rm l}+M_{\rm b}+M_{\rm s}+M_{\rm r}$, and the standing mass of the plant is $M_{\rm t}=M_{\rm a}+M_{\rm h}$. A summary of all variables, units, and definitions is given in Table \ref{tab:definitions}, with further details on the parameter values applied given in Tables \ref{tab:params_core}--\ref{tab:params_hyper}.

Applying a standard approach, the amount of biomass growth on the plant, $\frac{{\rm d}B}{{\rm d}t}$, is given by the difference between income (photosynthesis) and losses (respiration and turnover)\citep{Makela-1997, Thornley-2000}:
\begin{equation}\label{eq:dbdt}
\frac{{\rm d}B}{{\rm d}t}
= \alpha_{\rm bio} \,
\alpha_{\rm y}
\bigg( A_{\rm l} \, \bar{p}(E) -
\, \sum_{i = {\rm l}, {\rm b}, {\rm s}, {\rm r}}{M_i \, r_i}\bigg)
- \sum_{i = {\rm l}, {\rm b}, {\rm s},  {\rm r}}{M_i \, k_i}.
\end{equation}
Photosynthesis is the product of the average photosynthetic rate per unit leaf area, $\bar{p}(E)$, and total leaf area, $A_{\rm l}$. We assume that $\bar{p}$ increases with canopy openness $E$, as per a standard light-response curve (see SI for details), and respiration and turnover rates of different tissues are constants that potentially differ with traits. The constants $\alpha_{\rm y}$ and $\alpha_{\rm bio}$ account for the yield (fraction of C fixed in biomass) and the conversion of CO$_2$ into units of biomass, respectively. While the {\plant} model can easily accommodate competitive shading via influences on $E$, in this analysis we grow individual plants under a fixed light environment so that we can better understand trait- and size-related effects. Many vegetation models also use a more-detailed physiological model for calculating $\bar{p}$ and $r_{\rm i}$ -- e.g. as functions of temperature -- though such detail will not qualitatively alter model behaviour.

\begin{table}[!hb]
 \caption{Variable definitions.}
\centering
\begin{tabular}{p{0.5cm}p{1.75cm}p{5cm}}
\toprule
Symbol & Unit & Description \\
\midrule
\multicolumn{3}{l}{\textbf{Traits (constant through ontogeny)}}\\
$\omega$ & kg & Seed mass ({\seed})  \\
$H_{{\rm mat}}$ & m & Height at maturation ({\hmat})\\
$\nu$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Leaf nitrogen per unit leaf area ({\nitrogen})  \\
$\phi$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Leaf mass per area ({\lma}) \\
$\rho$ & $\mathrm{kg}\,\mathrm{m}^{-3}$ & Wood density ({\wood})\\
\multicolumn{3}{l}{\textbf{State or rate variables (may vary through ontogeny)}} \\
$H$ & m & Height of a plant\\
$D$ & m & Stem diameter\\
$M_i$ & kg & Mass of tissue type $i$ retained on plant \\
$A_i$ & m$^2$ & Surface or cross-section are of tissue type $i$\\
$\frac{{\rm d}B}{{\rm d}t}$ & kg yr$^{-1}$& Biomass growth rate\\
$p,\bar{p}$ & mol yr$^{-1}$ m$^{-2}$ & Photosynthetic rate per unit area \\
$r_i$ & mol yr$^{-1}$ kg$^{-1}$  & Respiration rate per unit mass of tissue type $i$ \\
$k_i$ & yr$^{-1}$ & Turnover rate for tissue type $i$ \\
\multicolumn{3}{l}{\textbf{Environmental variables (fixed)}} \\
$E$ & & Canopy openness\\
\multicolumn{3}{l}{\textbf{Other parameters (constant throughout)}} \\
$\alpha_{\rm y}$ &  & Yield, fraction of carbon fixed into biomass\\
$\alpha_{\rm bio}$  & $\mathrm{kg}\,\mathrm{mol}^{-1}$ & Biomass per mol carbon \\
$\eta_c$ & & Crown-shape parameter\\
$\theta$ &  & Sapwood area per unit leaf area\\
$\alpha_{\rm l1}$ & m & Height of plant with leaf area of 1m$^2$ \\
$\alpha_{\rm r1}$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Root mass per unit leaf area \\
$\alpha_{\rm b1}$ &  & Ratio of bark area to sapwood area\\
\bottomrule
\end{tabular}
\addtabletext{For $M_i$, $r_i$, $k_i$ and \(A_i\), subscripts refer to: \({\rm l}\) = leaves, \({\rm b}\) = bark, \({\rm s}\) = sapwood,  \({\rm h}\) = heartwood, \({\rm r}\) = roots, \({\rm st}\) = total stem, \({\rm a}\) =  alive tissue.

@@

\matmethods{
\vspace{-1em}
